---
title: 'Workshop 6: Neural networks'
output: html_document
---

In this workshop we'll be looking at how to predict a measurement (numeric) variable using other data with neural networks.

We'll be using the same data as Workshop 4 and 5 on computer performance. Recall that we are interested in predicting the target `usr` as a function of the other variables.

We start by loading packages and the data, splitting the data as we did in Workshop 4. You may have to install the `nnet` package if you haven't already.

```{r, include=FALSE}
library(tidyverse)
library(rsample)
library(yardstick)
library(nnet)
library(parsnip)
library(recipes)

comp <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/comp.csv")

set.seed(1234)
split <- initial_split(comp, prop=3/4)
comp_train <- training(split)
comp_test  <- testing(split)
```

## Model fitting

Let's start by fitting a neural network model to the `comp_train` data using all predictors.

```{r}
set.seed(1945)
model1.nn <- nnet(usr ~ ., data = comp_train, linout=TRUE, size=11)
```

Recall that with the `nnet` function we need to specify `linout=TRUE` as we are predicting a numeric variable (by default the output of `nnet` will be a probability between 0 and 1 which isn't much use as the `usr` variable ranges from 60 to 99). We also specify `size=11` - this is the number of nodes in the hidden layer, which we've chosen to be the number of effective predictors (21) divided by 2, rounded up - typically a useful starting point.

We have used `set.seed` here as the weights in the neural network are assigned random initial values, so this gives us some reproducibility.

### Try yourself

1. Get predictions from `model1.nn` for `comp_test` into a data frame and use this to compute the MSE (or RMSE - your choice).

2. Try adding `decay=0.01` to the `nnet` function to get a new model `model2.nn`. This tunes the decay rate of weights during the back-propogation algorithm. Check how well this does. Remember to use the same `set.seed` to ensure the weights are initialised the same way.

3. Try fitting the above models instead using `parsnip`. The model specification should be `mlp(mode="regression", engine="nnet", hidden_units=11, penalty=0.01)`. Here, `mlp` is short for "multilayer perceptron". Note that you don't need the `linout` parameter - this is automatically set based on the `mode`.

## The need for scaling

You should find that you get very poor predictions compared to linear models, trees or forests. This is because the neural network is exponentiating the linear combination of predictors in order to decide whether the hidden nodes should be activated or not. But, our predictors are on very different scales and some of them are very large:

```{r}
comp %>% summarise(across(everything(), max))
```

If we exponentiate 2.4 million, we get (essentially) infinity. Thus, hidden nodes are turned on or off directly based on the initial random weights, and there is no way to change them as any change in the weights will still give something like infinity (or minus infinity).

To deal with this we need to scale the data down to something small. A typical way to do this is to scale down so that the mean of each predictor is 0 and the standard deviation is 1. We scale linearly so we don't affect the shape of the data by subtracing the mean of each column and dividing by the standard deviation. Ofcourse, we have to make sure we scale the data the same way for both the training and test data.

One way to do this is to compute the necessary scaling factors (mean and sd) on the training data, and then applying those via the `scale` function to both training and test data. Using base R this would be:

```{r}
comp_train_mean <- comp_train %>% select(-usr) %>% apply(2, mean)
comp_train_sd   <- comp_train %>% select(-usr) %>% apply(2, sd)
comp_train_scale <- bind_cols(comp_train %>% select(usr),
                              comp_train %>% select(-usr) %>%
                                scale(center=comp_train_mean, scale=comp_train_sd) %>%
                                as.data.frame())
```

This is a little messy, but it works. We're using the `apply` function to compute column mean and sd and then using the `scale` function to do the scaling, but this returns a matrix, not a data.frame, so we have to transfer back again then bind the columns back together.

We could do this by staying within the tidyverse with dplyr - using the usual trick of pivoting longer first. We start by computing the scaling information:

```{r}
comp_train_scaling <- comp_train %>%
  pivot_longer(-usr, names_to='name', values_to='value') %>%
  group_by(name) %>%
  summarise(mean = mean(value), sd=sd(value))
```

Then we apply to the training data by adding a rowid (so we can pivot back wide again), 
joining to our scaling information, mutating to get the new value and then pivoting back after
removing the unused mean and sd columns. The same code can then be applied to `comp_test`

```{r}
comp_train_scale <- comp_train %>% rowid_to_column() %>%
  pivot_longer(-c(usr, rowid), names_to='name', values_to='value') %>%
  left_join(comp_train_scaling) %>%
  group_by(name) %>%
  mutate(value = (value - mean)/sd, .keep='unused') %>%
  pivot_wider(names_from=name, values_from=value) %>%
  select(-rowid)
```

### Try yourself

1. Use the above code to scale the `comp_test` data as well into `comp_test_scale`.

2. Fit another neural net this time using the scaled data. It should perform rather better than the unscaled ones.

## Tidying up the process of processing

The scaling of the data above, whether we use the `scale` command directly or whether we stay within the tidyverse is rather cumbersome - and easy to mess up. e.g. you might forget to apply the mean and sd from the training data to the test data, and instead accidentally recompute on the test data so the two scalings don't match.

Ofcourse, there's a package for that! We can utilise the `recipe` package within the `tidymodels` framework to handle this.

To perform scaling we have a simple recipe:

```{r}
comp_rec <- recipe(usr ~ ., data=comp_train) %>%
  step_normalize(all_numeric_predictors())

comp_rec
```
We give the recipe the formula and data (so it knows the outcome variable and other potential predictors) and then add the `step_normalize` function to normalise all numeric predictors.

The recipe just describes what we want to do: It doesn't actually do it! To perform the operation there are two more steps needed: We first must `prep`, which will estimate the parameters for scaling:

```{r}
trained_rec <- prep(comp_rec, training=comp_train)
trained_rec
```

Finally we need to `bake` our recipe using both our training and testing data. This will perform the needed transformations and return our datasets to us:

```{r}
comp_train_baked <- trained_rec %>% bake(comp_train)
comp_test_baked  <- trained_rec %>% bake(comp_test)
```

We can check these are the same as we got before with our manual setup by using `anti_join` - we should get no rows if they are the same:

```{r}
comp_train_baked %>% anti_join(comp_train_scale)
```

So steps to fit our model would be:

1. Use `recipe` to setup and process the data, and `prep` it on the training data.
2. Create a model specification (e.g. using `linear_reg` or `mlp`).
3. Fit the model using the `bake`'d training data.
4. Predict using model using the `bake`'d test data, and bind it to the test data.
5. Evaluate the model `metrics`.

In one code block this is:

```{r}
# Create our recipe:
comp_rec <-  recipe(usr ~ ., data = comp_train) %>%
  step_normalize(all_numeric_predictors()) %>%
  prep(training = comp_train)

# Create the model
nn_spec <- mlp(mode = "regression", hidden_units = 11, penalty = 0.01)

# Train the model
nn_fit <- nn_spec %>%
  fit(usr ~ ., data = bake(comp_rec, comp_train))

# Apply to test set and evaluate
nn_pred <- comp_test %>%
  bind_cols(
    nn_fit %>% predict(new_data = bake(comp_rec, comp_test))
  )

# Evaluate the model
nn_pred %>% metrics(usr, .pred)
```

### Try yourself

1. Run the code above a few times. You should notice that you get quite different metrics from run to run. The reason for that is the model didn't converge in the default number of steps.

2. Try altering the above to give it more steps by setting the `epochs` parameter. By default this is set to 100, so try 10000 to give it 100 more times. You should still see different metrics, but they shouldn't vary as much.

3. Why are you still getting different metrics from the neural network from run to run?

## Prediction competition: Bank data

For this next exercise we'll look at predicting whether a customer will reject a bank.

The (simulated) data are read in here:

```{r}
bank <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/bank-train.csv")
bank
```

The goal is to build a prediction model for the `rejection` variable, the probability of the customer rejecting the bank:

 - You should split the data into training and test sets.
 - The variables with the exception of `queue` are already pre-scaled, so you shouldn't need to use a `recipe`.
 - You should train a range of prediction models on the training set.
 - Choose the best model using predictive performance (RMSE) on your test set.
 
Once you have the best model, re-train it on all the data and use the model to obtain predictions
for this dataset:

```{r}
bank_predict_me <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/bank-predict-me.csv")
```

Notice that this second dataset does not have the known `rejection` variable - your goal is to provide the best predictions you can for this unknown.

Once you have your predicted rejection, you can write it to a csv file (see `write_csv`) and email it to: j.c.marshall@massey.ac.nz

Have fun!
