---
title: 'Workshop 5: Trees and Forests'
output: html_document
---

In this workshop we'll be looking at how to predict a measurement (numeric) variable using other data with regression trees and forests.

We'll be using the same data as Workshop 4 on computer performance. Recall that we are interested in predicting the target `usr` as a function of the other variables.

We start by loading packages and the data, splitting the data as we did in Workshop 4. You may have to install the `rpart`, `randomForest` and `parsnip` packages first.

```{r, include=FALSE}
library(tidyverse)
library(rsample)
library(broom)
library(yardstick)
library(rpart)
library(randomForest)
library(parsnip)

comp <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/comp.csv")

set.seed(1234)
split <- initial_split(comp, prop=3/4)
comp_train <- training(split)
comp_test  <- testing(split)
```

## Model fitting

Let's start by fitting a regression tree model to the `comp_train` data using all predictors. We then plot the tree with `plot`.

```{r}
model1.rp <- rpart(usr ~ ., data = comp_train)
plot(model1.rp, margin=0.1)
text(model1.rp)
```

The above tree is fit with the default value of the complexity parameter `cp`. We could try fitting a more complex tree by altering this:

```{r}
model2.rp <- rpart(usr ~ ., data=comp_train, cp=0.001)
```

### Try yourself

Plot and label this second tree. You should see you get a much more complex tree!

We can get a handle on an appropriate level for `cp` by utilising the built-in cross validation that is performed by `rpart` via the `printcp` function:

```{r}
cp_table <- printcp(model2.rp)
```

What we want to do is pick the largest value of `cp` (to give the smallest tree) such that the error in the tree is about the same as the 'optimal' tree. We typically choose the value of `cp` so that it's within one standard error from the minimum value:

```{r}
cp_table %>% as_tibble() %>%
  mutate(limit = min(xerror + xstd)) %>%
  filter(xerror <= limit)
```

So we should choose `cp = 0.003` or thereabouts. NOTE: Depending on random number seeds, you may end up with a different table - that's OK, we'll just rock on with `cp = 0.003`.

We can also produce a visual indication of this which is much easier to use:

```{r}
plotcp(model2.rp)
```

Again, about 0.003 seems to be where we no longer improve.

### Try yourself

Refit the tree model using `cp=0.003` and take a look at the corresponding tree.

## Predictive performance

To measure our predictive performance we do the same thing as before. Unfortunately, `rpart` is not included in `broom`, so there isn't a nice tidy way of doing this using the same framework. We could use the `predict` command. e.g. for `model1.rp` we'd have:

```{r}
predict(model1.rp, newdata=comp_test)
```

This returns a vector. We can add it to the original data via `bind_cols`, but this is a little fragile, as we're assuming that `predict` will return a row for everything we provide, and this may not be the case (e.g. if the test set had missing values perhaps).

```{r}
comp_pred1 <- comp_test %>%
  bind_cols(.fitted = predict(model1.rp, newdata=comp_test)) %>%
  select(usr, .fitted, everything())
comp_pred1
```

Now we can compute MSE in the same way we did in workshop 4:

```{r}
comp_pred1 %>% summarise(MSE = mean((.fitted - usr)^2))
```

### Try yourself

1. Compute the MSE for `model2.rp` and `model3.rp` in the same way. Which model performs best?

2. Try using `yardstick` to compute the `rmse` and `mae` for these models.

## Dealing with missingness

One thing that regression trees are particularly useful for are where we have some data missing.

Let's take our training set and change it so we have some missing data in `vflt` (recall this was the variable used in the first split) [^na_real].

```{r}
comp_train.miss <- comp_train %>%
  mutate(vflt = if_else(row_number() <= 5, NA_real_, vflt))
comp_train.miss %>% select(vflt, everything())
```

### Try yourself

1. Fit a regression tree to the `comp_train.miss` data using `cp=0.003` as above.
2. Plot the tree. You should find that `vflt` is still the first split.
3. Take a look at the output for `summary()` on this tree. You should see it is using a *surrogate split*. Which variable is being used for the split if `vflt` is missing?

## Moving from trees to forests

Instead of fitting a single tree to the data, we could instead try lots of trees: a forest. A random forest is where we re-sample the training data repeatedly (using the bootstrap) and for each resample we fit a tree (using a random subset of the predictors rather than all of them). We then average our predictions across all the trees. It's an example of an **ensemble** method, as we're essentially averaging over many models.

These two things allow us to explore a range of potentially different trees, allowing some variables to be used that might not otherwise be in a single tree, and potentially giving a better overall prediction.

The `randomForest` package can do this. The `importance` function then gives us an idea of which variables are important. We can plot this with `varImpPlot()`[^vip].

```{r}
model.rf <- randomForest(usr ~ ., data=comp_train)
importance(model.rf)
varImpPlot(model.rf)
```

As expected, the `vflt` and `pflt` variables are most important.

### Try yourself

1. The `predict()` function works in the same way as it does for `rpart`. Produce predictions from the random forest model.

2. Compute MSE for the random forest model compare with the decision trees?

3. There are two parameters that might affect how well the forest fits: `mtry` which is the number of predictors to use for building each tree (by default it is set to number of predictors divided by 3) and `nodesize` which defines the minimum size of terminal nodes, which defaults to 5. Try changing each of these and seeing how they change model performance.

## Making things a bit tidier with parsnip

The commands required to fit a model with `rpart` or `randomForest`, in particular to do prediction are a little messy, and unfortunately `broom` doesn't sweep things tidy for tree based models[^broomstick].

The `tidymodels` framework is a way to address this for `rpart`, `randomForest` and a bunch more models. The `broom` and `yardstick` packages are part of this. The goal is to provide a uniform interface for both fitting models ( with `lm`, `rpart` and `randomForest`, these have so far been similar) and for obtaining predictions or model performance. The `parsnip` package is what provides a common model-fitting interface[^parsnip].

```{r}
tree_spec <- decision_tree(mode = "regression",
                            cost_complexity = 0.01)

tree_fit <- tree_spec %>%
  fit(usr ~ ., data = comp_train)

tree_pred <- comp_test %>%
  bind_cols(predict(tree_fit, new_data = comp_test)) %>%
  select(usr, .pred, everything())

tree_pred %>% metrics(usr, .pred)
```

Note the process here:

1. We first specify the model type (a decision tree model in regression mode).
2. We then fit the model via `fit()`, specifying the formula and training data.
3. We then predict the model via `predict()`, specifying the test data.
4. We then summarise the fit using `metrics()` (or `rmse`, `mae` etc).

### Try yourself

1. Modify the above to provide fits with different cost complexity (`cp`).

2. If you want to get the actual `rpart` model object out of the `tree_fit` object you can use `extract_fit_engine(tree_fit)`. Try this to do the `plotcp` plot.

3. Change the above so it uses a linear regression model instead: You need only change the `decision_tree` command to `linear_reg`.

4. Change the above so it uses a random forest instead by using `rand_forest(mode = "regression", engine = "randomForest")`. The `engine` argument is because the default engine is `ranger`. You can use that if you like too! You might want to try exploring the `mtry` and `min_n` arguments.


[^na_real]: Note the use of `NA_real_` here. This is because the `if_else` function (from `dplyr`) is opinionated! It hates it when you provide two different types of variable to the true or false parameters. The default `NA` is of type logical, so we'd be mixing a logical type and a numeric type if we tried using that. Alternatively, you could use `ifelse()` which doesn't have such strong opinions!

[^broomstick]: There is a package `broomstick` (neat name for a tree cleaning broom...) but this is not yet on CRAN, and still seems to be a work in progress.

[^parsnip]: The package name here is a homage to the `caret` (short for classification and regression training) package which provides the same idea, albeit with a very different implementation.

[^vip]: The `vip` package contains a the `vi` function for a tidy version of `importance` and `vip` for a `ggplot` version of `varImpPlot`.