---
title: 'Workshop 8: Regression classifiers'
output: html_document
---

In this workshop we'll be looking at classifiers based on the generalised linear model:

- Logistic regression
- Multinomial regression

## Logistic regression for Horse Colic data

This Exercise uses the data on horses with colic. Colic can be a serious
illness for horses and can result in death. Physiological data on 368
horses with Colic were recorded, along with whether surgery was
required, with some of the horses dying and some surviving.

The goal is to train a logistic regression classifier in order to
classify the horses into whether they're likely to survive or die based
on the physiological parameters measured.

We'll start by loading packages and reading in the data, and then converting a bunch of the columns to factors:

```{r, message=FALSE}
library(tidyverse)
library(rsample)
library(skimr)
library(discrim)
library(yardstick)
library(recipes)
step <- stats::step # override recipes::step() with stats::step()

horse.orig <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/horse2.csv")

horse <- horse.orig %>%
  mutate(across(-c(RectalTemp, Pulse, RespRate, PckCellVol, TotlProtein), as_factor))

set.seed(2013)

split <- initial_split(horse, prop=0.75)
horse.train <- training(split)
horse.test  <- testing(split)
```

### Try yourself

1. Use `logistic_reg()` with the `glm` engine to build a logistic model for `Died` based on all the other physiological parameters.

2. Take a look at the model summary - recall that you can do this with `tidy()` or alternatively `extract_fit_engine()` and then `summary()`.

3. Compute the odds ratio for the `Surgery` predictor, where a 1 specifies that the horse had surgery. Does having surgery increase or decrease the odds of the horse dying?

4. Perform a prediction on the test data set for whether the horses will die or not, and assess the misclassification rate.

5. Try using the `step` command to derive a simpler model for the data, and check how the simpler model performs as a classifier. *NOTE: You'll have to re-fit the model directly using `glm` before running `step`. Once done, you can feed the final model formula into `fit()` to get back into tidymodel land!*

## Classifying the Italian wine data

We'll use multinomial regression on the Italian wine data to examine the effects of data scaling on classification results.

We'll start by loading the wine test and training data in:

```{r, message=FALSE}
wine.train <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/wine-train.csv") %>%
  mutate(Cultivar = factor(Cultivar))
wine.test  <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/wine-test.csv") %>%
  mutate(Cultivar = factor(Cultivar))
```

### Try yourself

1. Use `multinom_reg()` to fit a multinomial regression model on all the data. Remember that the number of iterations (epochs) is important to ensuring it converges.

2. Evaluate the performance of classification on the test set by computing accuracy and a confusion table.

3. Now create a `recipe` to scale the numeric predictors, and refit the multinomial regression model using the scaled data.

4. Evaluate the performance of the classifiers before and after scaling. Which does better?

5. You might want to investigate running `step()` on these multinomial models to simplify them. Again, this has to be done outside the `tidymodels` framework.
