---
title: 'Workshop 10: Clustering'
output: html_document
---

In this workshop we'll be looking at clustering data. Clustering is an unsupervised
learning technique - there is no known truth, and we're instead trying to group the
observations into a fairly small number of clusters such that all observations in a
clusters are closely related, while observations from different clusters differ.

The techniques we'll cover include:

- Hierarchical clustering
- k-means clustering
- Partitioning around Medioids
- Silhouttes

## The 2020 New Zealand General Election

In this exercise we'll look at the results from the 2020 New Zealand general election,
and look at how the 72 electorates (regions) cluster together based on the party
and candidate votes, or both combined.

In New Zealand general elections there are 65 'General' electorates (regions) and 7
Maori electorates. Each voters is assigned to one of these electorates, and has two
votes: they vote for the political party they wish to represent them in parliament,
and in addition vote for the person (candidate) they wish to represent them in their
electorate.

Generally, the candidate they vote for in an electorate is also in a political party,
but there is no need for an individual to vote for the same party across both their
votes - indeed, vote splitting, where one votes for a person from a party other than
the one you vote for is common and encouraged in some electorates.

For more information about voting in New Zealand see here:

https://elections.nz/democracy-in-nz/what-is-new-zealands-system-of-government/

We'll be usig data from the 2020 general election, which we read in below.

```{r, message=FALSE}
library(tidyverse)
library(broom)
library(ggdendro)
library(cluster)
library(recipes)

party  <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nzelect/2020_party.csv")
candidate <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/nzelect/2020_candidate.csv")
```

### Try yourself

1. Take a look at the `party` data. You should notice you have the percentage vote for each party (in separate columns) by the electorate (in rows). Summarise how each party does on average across the electorates.

2. Produce a long-form dataset `party_long` by pivoting the vote columns into `Party` and `PartyVote` columns. Use this to produce boxplots of the `PartyVote` by `Party` to show the variation across the electorates.

3. Produce a long-form version of the `candidate` data, and combine both into a `all_long` dataset. Use this to do a scatterplot of the person vote versus the party vote, coloured by Party.


We'll next look at clustering the `party` data using hierarchical clustering:

```{r}
party.hc1 <- party %>%
  column_to_rownames("Electorate") %>%
  dist() %>%
  hclust(method='single')

ggdendrogram(party.hc1)
```

### Try yourself

1. Try performing hierarchical clustering with average or complete linkage.

2. Which electorates tend to cluster together?

Let's now look into the clustering a little more, by using `cutree` to break things down into (say) 5
clusters:

```{r}
party.cl1 <- party %>% 
  mutate(Cluster = party.hc1 %>% cutree(k=5))

party.cl1 %>%
  group_by(Cluster) %>%
  summarise(across(-Electorate, mean), size=n())
```

We can see that cluster 5 (The Maori electorates) all teend to have a much higher vote for the Maori and Other parties and a lower vote for National, while cluster 1 tends to vote more for National. Cluster 3 are South Auckland electorates that vote very strongly for Labour, while Cluster 2 and 4 vote more strongly for the Greens (two Wellington electorates and Dunedin).

### Try yourself

1. Do the other linkage methods result in the same clustering?

2. Do the same analysis of the `candidate` vote. How do the electorates cluster? Choose a suitable number of clusters based on the dendogram and then look into which electorates fall in each cluster.

We'll now combine the two datasets so we can look at them together:

```{r}
all <- party %>% rename_with(~ paste("Party", ., sep=":"), .cols=-Electorate) %>%
  left_join(
    candidate %>% rename_with(~ paste("Cand", ., sep=":"), .cols=-Electorate),
    by = 'Electorate'
  )
```

### Try yourself

Using the combined voting data `all`, perform hierarchical clustering and investigate which
electorates tend to vote similarly when both party and candidate votes are considered.


## Clustering synthetic data

In this exercise we have 9 numerical measurements (labelled A through I)
from a synthetic data set. Our goal is to group the observations with
common characteristics together.

```{r, message=FALSE}
synth <- read_csv("https://www.massey.ac.nz/~jcmarsha/data/syn-cluster.csv")
```

### Try yourself

1. Start by using hierarchical clustering to look at this data. How many clusters look sensible?


We'll next try using k-means clustering, with the following code fitting k-means for k=1 through 10:

```{r}
set.seed(1234)
km_all <- tibble(k = 1:10) %>%
  mutate(
    model = map(k, ~ kmeans(synth, centers = ., nstart = 20)),
    tidied = map(model, glance)
  ) %>%
  unnest(tidied)
km_all
```

### Try yourself

1. Produce a plot of the within sum of squares against k. How many are clusters do you think
are optimal?

2. Check the scales on each of the variables in `synth`. You should notice they are different!

We can rescale the data using a `recipe` as usual:

```{r}
scaling <- recipe(synth) %>%
  step_normalize(everything()) %>%
  prep(synth)

synth.scale <- bake(scaling, synth)
```

### Try yourself

1. Repeat the clustering procedure. How many clusters make sense now?


You should notice that 6 clusters seems sensible. We can confirm this by looking at the silhouettes of 6 and 7 clusters. If 6 clusters is best, then by asking for 7 clusters we will likely split one of the clusters artificially, which should result in at least two clusters with poor silhouette widths.

We can extract the silhouette information by re-fitting using `pam`:

```{r}
clust6 <- pam(synth, k=6)
clust6.sil <- clust6 %>% pluck('silinfo', 'widths') %>% as_tibble()
```

### Try yourself

1. Create boxplots of the silhouette width by cluster using `clust6.sil`.

2. Repeat this for k=7. You might wish to combine the plots into one by using `bind_rows()` to 
produce a single plot.

3. Look at the results of your preferred clustering by using `augment()` from `broom`
to append the cluster number to the data. With these data, just doing boxplots for each variable
by the cluster will be enough to see how clear things are!
