# Data Acquisition and Processing

## Obtaining Data

Several datasets are provided for your study of data mining. They come from
a variety of fields, including business and commerce, ecology, finance
and genetics. As mentioned previously, all of the datasets that we shall
study are large, but none are massive. Any computer less than about ten
years old should be able to handle them. Instructions for acquiring
these data appear as required throughout these notes.

Quite possibly you will have (either now or at some point in the future)
data that of your own that you wish to analyze in R. R can import data
that is stored in a wide variety of formats. We discuss data importation
in the next section.

If you are looking for other data sources for data mining, then there
are a wide variety available on the Web. Here we mention just three. The
bioconductor project, `www.bioconductor.org`, is a valuable resource for
anyone interested in bioinformatic problems. Bioconductor is designed
for use with R. If economics and finance is your thing then the Yahoo
site `finance.yahoo.com` provides free access to a range of financial
data. Finally, those who fancy a challenge might like to visit the
Kaggle website `www.kaggle.com`. Kaggle runs a series of classification
and prediction competitions, some with quite large prizes available!

## Loading Data into R{#sec:dataload}

One method of reading data into R is using `read.table`. This
accesses data stored in tabulated text format and loads it
as a data frame.

By default `read.table` assumes that the columns of the target text file
are separated by white space, and that the first row of the text file
comprises data items rather than column (variable) names. These options
can be altered by setting the `sep` and `header` arguments of the
function `read.table`. The first of these specifies the column separator
character, which should be appear between speech marks in the function
call. The second takes logical arguments. It is `FALSE` by default (i.e. the
first row of the data is not a header), but can be set to `TRUE`.

This might be easier to see through a specific example, so consider the
following command:

```{r, eval=FALSE}
fires <- read.table(file="forestfires.csv", sep=",", header=TRUE)
```

This reads data from the file `forestfires.csv`, a comma separated
values text file, and assigns the result to a data frame named `fires`.
The separator is set to a comma, and the first row of `forestfires.csv`
is interpreted as variable names. Exactly the same effect can be
achieved using the function `read.csv`, which is the same as
`read.table` except for the default settings for `sep`, `header` etc.
Hence

```{r, eval=FALSE}
fires <- read.csv(file="forestfires.csv")
```

produces exactly the same result as the code above.

The facility to load data from text files in any (sane) format provides
the means for importing data from many statistical, spreadsheet and
database packages. Most such packages can save files in a plain text
format: for instance, Excel has the option to save spreadsheets as
`.csv` (comma separated valued) files. These can then be read into R. As
an alternative, the R package `Foreign` includes some functions for
direct importation into R from native file types for some packages
(including Minitab, S, SAS, SPSS, Stata, and Systat)[^otherpackages].

## Organizing the Data in R

The common format for datasets is as a matrix, with records (also known
as observations, cases, or units) as rows and variables (also known as
attributes) as columns. The information in each column may be numerical
(in the case of quantitative variables) or categorical (in the case of
factors[^factors]). For example, the first six rows of the goats data (as
displayed in R) are

    Treatment       Wt.ini  Wt.gain
    1  standard     21       5
    2  standard     24       3
    3  standard     21       8
    4  standard     22       7
    5  standard     23       6
    6  standard     26       4

where the row numbers identify individual goats, and the columns
correspond to the variables in the obvious manner. The number of records
is usually denoted $n$, and the number of variables $p$.

The name for a data structure of this type is a *data frame*.
As we just saw, this is the type of object
that is created when you employ the `read.table` command. However, on
occasions you may find that your dataset requires some manipulation in R
to get it into our preferred format. We will illustrate this through a
specific example.

::: {.example}
**Bad Goats**

Some researchers (fortunately a minority nowadays) like to organize data
so that columns refer to specific groups. For example, one could arrange
the goats data in a text file with four columns, with the first pair
being initial weight and weight gain for the standard treatment group,
and the second pair being initial weight and weight gain for the
intensive drenching group. The text file (which we will call
`bad-goats.txt` below) would then look something like this

    "iniS"  "gainS" "iniI"  "gainI"
    21      5       18      9
    24      3       18      8
    21      8       19      8
    22      7       19      8
    ....

where the column names specify the variable and the group in the natural
way.

Suppose that you are presented with this text file? How can you
manipulate it to obtain the nice 'records as rows' format? The following
R snippet will do the trick.

```{r}
bad.goats <- read.table(file="data/bad-goats.txt", header=TRUE)
head(bad.goats)
grp.size <- nrow(bad.goats)
grp.size
Wt.ini <- c(bad.goats$iniS,bad.goats$iniI)
Wt.ini
Wt.gain <- c(bad.goats$gainS,bad.goats$gainI)
Wt.gain
Treatment <- factor(c(rep("standard",grp.size),rep("intensive",grp.size)))
Treatment
goats <- data.frame(Treatment,Wt.ini,Wt.gain)
head(goats)
```

OK, don't panic! I know there may seem to be a lot going on there, but
it really isn't very complicated.

-   The first command,
    `bad.goats <- read.table(file="data/bad-goats.txt", header=TRUE)`, simply
    reads in the badly formatted data from the text file, and assigns it
    to a data frame `bad.goats`.

-   The next command, `head(bad.goats)`, prints out the first size rows
    of this data frame.

-   The function `nrow` returns the number of rows in a data frame or
    matrix, so `grp.size <- nrow(bad.goats)` stores the number of rows
    in our data frame `bad.goats` as the object `grp.size`. Now if you
    think about it, the number of rows in this data frame is just the
    number of recorded weights for each goat in each group, and is hence
    the group size. Of course, you might have remembered that this was
    20, but the code demonstrates what to do if you don't.

-   Recall that the syntax `df$variable` refers to a `variable` within a
    data frame `df`. Hence `bad.goats$iniS` is the variable `iniS`
    within the data frame `bad.goats`. It follows that

    ```{r, eval=FALSE}
    Wt.ini <- c(bad.goats$iniS, bad.goats$iniI)
    ```

    creates a vector `Wt.ini` which is the concatenation of all the
    initial weights, with those from the standard treatment group
    included first. Similarly,

    ```{r, eval=FALSE}
    Wt.gain <- c(bad.goats$gainS, bad.goats$gainI)
    ```

    creates the variable `Wt.gain` by concatenating the vectors of
    weight gains from the standard and treatment groups respectively.

-   The command `c(rep("standard", grp.size)` creates a character vector
    containing the word 'standard' replicated 20 (i.e. `grp.size`)
    times. Therefore

    ```{r, eval=FALSE}
    factor(c(rep("standard",grp.size),rep("intensive",grp.size)))
    ```

    produces a *factor* with two levels, `standard` and `intensive`,
    with the first 20 elements being the former and the final 20
    elements the latter.

-   Finally, the `data.frame` command is used to produce a new data
    frame, in precisely the desired format.
:::

We have one final comment on data organization before we move on.
Remember that when you use `read.table`, R will assume by default that
any column of numbers corresponds to a quantitative variable. Consider
the following example.

```{r, error=TRUE}
fossils <- read.table(file="data/fossils.txt", header=TRUE)
fossils
sqrt(fossils$species)
fossils$species <- as.factor(fossils$species)
sqrt(fossils$species)
```

Here we are dealing with a (small) dataset comprising the bite sizes
measured on 27 fossils of prehistoric fish. These fish are classified by
species, denoted simply 1, 2, 3. When the data are first read in the
`species` variable is interpreted as numerical, and so you can perform
numeric operations (like taking a square root). Of course, the numbers
are merely labels here. We can turn the numeric variable into a factor
using the `as.factor` command. Once R recognizes that `species` is a
factor (i.e. a categorical variable) it will no longer permit such
numerical operations.

## Missing Values{#sec:missing1}

Real datasets often contain missing values. The reasons are various. For
example, in socio-economic surveys some respondents are unwilling to
provide what they regard as quite personal information (e.g. gross
income, marital status). In scientific experiments one may fail to
record information from all the experimental units: perhaps someone
dropped a Petri dish, or one of the lab rats died. When missing values
are present we need to think carefully about what should be done about
them, and provide details of what was done in any reported analysis of
the data[^droppingout].

When you are presented with data in a text file, the presence of missing
values might be indicated in any of the following ways.

-   In a comma separate file (`.csv`) a missing value will often be
    indicated by an absence of any value between two consecutive commas.
    These missing values will be handled appropriately by R when using
    `read.csv` (or `read.table` with the right settings).

-   In a tab delimited text file there may simply be white space at a
    particular element. As before, the missing values will (usually) be
    correctly handled by the `read.table` command.

-   If row entries are separate by variable white space in a text file,
    the it can be difficult to handle missing values (unless they appear
    only in the last variable, when the `fill` option of `read.table`
    can be helpful). It is probably safest in this case to ask for the
    data in a better format!

-   Sometimes missing values are specified in text files by sequences
    like `9999` or `-9999`. This generally works OK, so long as the
    value 9999 (or whatever alternative is used) is not a plausible
    value for the variable in question! Nonetheless, one of the
    disadvantages of this representation is that it is possible that the
    presence of the missing values will go unnoticed, and the data will
    get analysed as if the 9999 entries are *bona fide* data. Note that
    it is possible to set the argument `na.strings` of `read.table` so
    that R will correctly interpret missing values specified in this
    manner.

-   A common alternative is to represent missing values in a text file
    by `NA` (for '**n**ot **a**vailable'). Of course, one has to be
    careful that the string NA does not have some credible meaning in a
    text field[^nasodium].

Regardless of the representation of missing values in the text file,
once that data are in R the missing values are denoted by `NA`. R
recognizes missing values as a special type of data, and has methods for
handling datasets with missing values when conducting certain kinds of
analysis. Nonetheless, in many cases the presence of one or more missing
values will prevent R from returning a numerical result. We will look at
ways of dealing with this problem in section \@ref{sec:missing}.

::: {.example}
**Husbands and Wives**

To illustrate handling of missing values we turn to a modestly sized
dataset on husbands and wives. (The data source is Hand, Daly, Lunn,
McConway and Ostrowski (1993), *A Handbook of Small Datasets*, Chapman
and Hall.) Each of the 199 records corresponds to a husband-wife pair,
for which the following variables are recorded.

  ------------------ ---------------------------------------------
  `H.Age`            Husband's current age (in years)
  `H.Ht`             Husband's height in millimetres
  `W.Age`            Wife's current age (in years)
  `W.Ht`             Wife's height in millimetres
  `H.Age.Marriage`   Husband's age (in years) when first married
  ------------------ ---------------------------------------------

The data contain a number of missing values, represented in the text
file of the data by the number 9999. These data are processed in the
following R snippet.

```{r}
husbands <- read.table(file="data/husbands.txt", header=TRUE)
str(husbands)
summary(husbands)
mean(husbands$W.Age)
husbands$W.Age[husbands$W.Age==9999] <- NA
mean(husbands$W.Age)
mean(husbands$W.Age, na.rm=TRUE)
husbands$H.Age.Marriage[husbands$H.Age.Marriage==9999] <- NA
mean(husbands$H.Age.Marriage, na.rm=TRUE)
```

Some points to note.

-   The command `str` provides a short summary of a data frame,
    including the variable type (in this example `int`, i.e. integer for
    all variables) and a listing of the first few elements for each
    variable.

-   The `summary` command applied to a data frame gives six figure
    summaries for all numerical variables. Note that `W.Age` and
    `H.Age.Marriage` have maximum values of 9999, corresponding to
    missing data.

-   If we carelessly forget to recode the missing values, then we will
    get nonsense in numerical values. For example, it would appear that
    the mean age of the wives is just a shade under 1500 years.

-   The syntax `husbands$W.Age[husbands$W.Age==9999] <- NA` selects all
    entries of the vector `W.Age` (in the data frame `husbands`) that
    are currently 9999 and replaces them with the missing value
    indicator `NA`.

-   Applying the function `mean` to a variable containing missing values
    will by default return `NA` as a result. However, if we set the
    optional argument `na.rm=T` (i.e. remove missing values) then we get
    the mean for the remaining data. In the case of wives' ages, that is
    just over 40.
:::

[^otherpackages]: R can also interact with database systems as well, although the
    process is a little more complex.

[^factors]: As noted previously, categorical variables in R are typically
    referred to as factors. The different categories themselves are
    termed the *levels* of the factor. Hence the factor `Treatment` in
    the goats dataset has two levels, `standard` and `intensive`
    distinguishing the drenching routine employed.

[^droppingout]: An important example occurs with data from clinical trials, where
    the statistical analysis is typically expected to account for
    *intention to treat*. This prevents a misleadingly statistical
    results in cases where subjects who are responding poorly have a
    disproportionate chance of dropping out of the trial.

[^nasodium]: You might run into problems if the data include a list of chemical
    elements, for example! The chemical symbol for sodium is Na, from
    the Latin word *natrium*.
