# Data Acquisition and Processing

## Obtaining Data

Several datasets are provided for your study of data mining. They come from
a variety of fields, including business and commerce, ecology, finance
and genetics. As mentioned previously, all of the datasets that we shall
study are large, but none are massive. Any computer less than about ten
years old should be able to handle them. Instructions for acquiring
these data appear as required throughout these notes.

Quite possibly you will have (either now or at some point in the future)
data that of your own that you wish to analyze in R. R can import data
that is stored in a wide variety of formats. We discuss data importation
in the next section.

If you are looking for other data sources for data mining, then there
are a wide variety available on the Web. Here we mention just three. The
bioconductor project, `www.bioconductor.org`, is a valuable resource for
anyone interested in bioinformatic problems. Bioconductor is designed
for use with R. If economics and finance is your thing then the Yahoo
site `finance.yahoo.com` provides free access to a range of financial
data. Finally, those who fancy a challenge might like to visit the
Kaggle website `www.kaggle.com`. Kaggle runs a series of classification
and prediction competitions, some with quite large prizes available!

## Loading Data into R{#sec:dataload}

We'll be mostly using the `readr` package for reading data into R. This is
part of the (https://tidyverse.org)[tidyverse] set of packages. In general
we'll be using the tidyverse throughout, at least where we can.

We'll start most of our code with:

```{r, message=FALSE}
library(tidyverse)
```

We'll tend to use CSV (comma separated value) files, for which the
`read_csv()` command is appropriate.

By default, this should work out of the box for just about all data
sets we use, but some poorly formatted data may need alternatives.

This might be easier to see through a specific example, so consider the
following command:

```{r}
fires <- read_csv(file="../data/forestfires.csv")
```

This reads data from the file `forestfires.csv`, a comma separated
values text file, which is located in the folder `../data` which means
"up to the parent folder and into the data folder". The result
is assigned to a data frame named `fires`.
Note that has also given us a message about how it has interpreted
the various columns in `forestfires.csv`, with some being interpreted
using `col_double()` (numeric, such as `X`, `Y`) and some being interpreted
using `col_character()` (character string, such as `month` and `day`).

The facility to load data CSV files provides
the means for importing data from many statistical, spreadsheet and
database packages. Most such packages can save files in a plain text
format: for instance, Excel has the option to save spreadsheets as
`.csv` (comma separated valued) files. These can then be read into R. As
an alternative, the R package `Foreign` includes some functions for
direct importation into R from native file types for some packages
(including Minitab, S, SAS, SPSS, Stata, and Systat)[^otherpackages], and
the `readxl` package can read directly from `.xlsx` Excel files.

## Organizing the Data in R

The common format for datasets is as an array, with records (also known
as observations, cases, or units) as rows and variables (also known as
attributes) as columns. The information in each column may be numerical
(in the case of quantitative variables) or categorical (in the case of
factors[^factors]). For example, the first six rows of the goats data (as
displayed in R) are

```{r, echo=FALSE, message=FALSE}
goats <- read_csv("../data/goats.csv")
head(goats)
```

where the row numbers identify individual goats, and the columns
correspond to the variables in the obvious manner. The number of records
is usually denoted $n$, and the number of variables $p$.

The name for a data structure of this type is a *tibble* or *data frame*[^tibble].
As we just saw, this is the type of object
that is created when you employ the `read_csv` command.

Sometimes you may wish to create a data set directly rather than reading it in
from a file. This is generally useful for small data sets, and can be done using
the `tibble`, `data.frame`, or `tribble` functions, as we demonstrate with the
following example.

::: {.example}
**Creating data in R**

Suppose we wish to create a small data set that looks something like this:

    x       y    group
    21      5        A
    24      3        A
    21      8        B
    22      7        B
    34      1        C
    18      8        C

where we have three columns `x`, `y` and `group` and six entries.

We can define this a number of ways as the below code illustrates.

```{r}
one <- data.frame(x = c(21, 24, 21, 22, 34, 18),
                  y = c(5, 3, 8, 7, 1, 8),
                  group = c("A", "A", "B", "B", "C", "C"))
two <- tibble(x = c(21, 24, 21, 22, 34, 18),
                  y = c(5, 3, 8, 7, 1, 8),
                  group = rep(c("A", "B", "C"), each=2))
three <- tribble(~x, ~y, ~group,
                 21, 5, "A",
                 24, 3, "A",
                 21, 8, "B",
                 22, 7, "B",
                 34, 1, "C",
                 18, 8, "C")
one
two
three
```

As you can see, each of these give the same data, but the method
of construction is a little different:

- The `data.frame` method returns a `data.frame` rather than a `tibble`,
so the printing is a little different.

- The `tibble` method is otherwise the same as the `data.frame` method,
though we have used the `rep` command here to repeat the `A`, `B`, `C` 
groupings each two times. This only slightly saves on typing!

- The `tribble` command (row-wise tibble) is the one that is most natural
for entering data that is tabular like this. The tilde character `~` is
used for specifying the names, and then everything else is just separated
by commas, and because it is filled row-wise rather than column-wise can
be written so that it maintains the natural ordering.
:::

We have one final comment on data organization before we move on.
Remember that when you use `read_csv`, R will assume by default that
any column of numbers corresponds to a quantitative variable. Consider
the following example.

```{r, error=TRUE}
fossils <- read_csv(file="../data/fossils.csv")
fossils
fossils %>% mutate(sqrt(species))
fixed_fossils <- fossils %>% mutate(species = as.factor(species))
fixed_fossils
fixed_fossils %>% mutate(sqrt(species))
```

Here we are dealing with a (small) dataset comprising the bite sizes
measured on 27 fossils of prehistoric fish. These fish are classified by
species, denoted simply 1, 2, 3. When the data are first read in the
`species` variable is interpreted as numerical, and so you can perform
numeric operations (like taking a square root). Of course, the numbers
are merely labels here. We can turn the numeric variable into a factor
using the `as.factor` command. Once R recognizes that `species` is a
factor (i.e. a categorical variable) it will no longer permit such
numerical operations.

The `visdat` package is useful for visualising a `data.frame` or `tibble`
to assess things like variable types.

::: {.example}
**Visualising variable types in the airquality dataset**

The `airquality` dataset, included with R in the `datasets` package, contains
153 observations on 6 variables (`Ozone`, `Solar.R`, `Wind`, `Temp`, `Month`,
and `Day`). We can visualise the structure of these using the `vis_dat()` function
from `visdat`:

```{r}
library(visdat)
vis_dat(airquality)
```

As can be seen, `vis_dat()` shows us that the first 5 columns are 'integer' type
(whole numbers) while the `Wind` column is 'numeric' (double precision real value).

We can also see where missing values are - in particular the `Ozone` variable has a bunch
of missing values.
:::

## Missing Values{#sec:missing1}

Real datasets often contain missing values. The reasons are various. For
example, in socio-economic surveys some respondents are unwilling to
provide what they regard as quite personal information (e.g. gross
income, marital status). In scientific experiments one may fail to
record information from all the experimental units: perhaps someone
dropped a Petri dish, or one of the lab rats died. When missing values
are present we need to think carefully about what should be done about
them, and provide details of what was done in any reported analysis of
the data[^droppingout].

When you are presented with data in a text file, the presence of missing
values might be indicated in any of the following ways.

-   In a comma separate file (`.csv`) a missing value will often be
    indicated by an absence of any value between two consecutive commas.
    These missing values will be handled appropriately by R when using
    `read_csv`.

-   Sometimes missing values are specified in text files by sequences
    like `9999` or `-9999`. This generally works OK, so long as the
    value 9999 (or whatever alternative is used) is not a plausible
    value for the variable in question! Nonetheless, one of the
    disadvantages of this representation is that it is possible that the
    presence of the missing values will go unnoticed, and the data will
    get analysed as if the 9999 entries are *bona fide* data. Note that
    it is possible to set the argument `na` of `read_csv` so
    that R will correctly interpret missing values specified in this
    manner.

-   A common alternative is to represent missing values in a text file
    by `NA` (for '**n**ot **a**vailable'). Of course, one has to be
    careful that the string `NA` does not have some credible meaning in a
    text field[^nasodium].

Regardless of the representation of missing values in the text file,
once that data are in R the missing values are denoted by `NA`. R
recognizes missing values as a special type of data, and has methods for
handling datasets with missing values when conducting certain kinds of
analysis. Nonetheless, in many cases the presence of one or more missing
values will prevent R from returning a numerical result. We will look at
ways of dealing with this problem in section \@ref{sec:missing}.

::: {.example}
**Husbands and Wives**

To illustrate handling of missing values we turn to a modestly sized
dataset on husbands and wives. (The data source is Hand, Daly, Lunn,
McConway and Ostrowski (1993), *A Handbook of Small Datasets*, Chapman
and Hall.) Each of the 199 records corresponds to a husband-wife pair,
for which the following variables are recorded.

  ------------------ ---------------------------------------------
  `H.Age`            Husband's current age (in years)
  `H.Ht`             Husband's height in millimetres
  `W.Age`            Wife's current age (in years)
  `W.Ht`             Wife's height in millimetres
  `H.Age.Marriage`   Husband's age (in years) when first married
  ------------------ ---------------------------------------------

The data contain a number of missing values, represented in the text
file of the data by the number 9999. These data are processed in the
following R snippet.

```{r, render = knitr::normal_print}
husbands <- read_csv(file="../data/husbands.csv")
glimpse(husbands)
skimr::skim(husbands)
husbands %>% summarise(mean(W.Age))
husb_fixed <- husbands %>%
  naniar::replace_with_na(replace = list(W.Age = 9999,
                                         H.Age.Marriage = 9999))
husb_fixed %>%
  summarise(mean(W.Age),
            mean(W.Age, na.rm=TRUE),
            mean(H.Age.Marriage, na.rm=TRUE))
```

Some points to note.

-   The command `glimpse` provides a short summary of a data frame,
    including the variable type (in this example `dbl`, i.e. integer for
    all variables) and a listing of the first few elements for each
    variable.

-   The `skimr::skim` command applied to a data frame gives summaries for
    all variables. Note that `W.Age` and
    `H.Age.Marriage` have maximum (`p100`) values of 9999, corresponding to
    missing data.

-   If we carelessly forget to recode the missing values, then we will
    get nonsense in numerical values. For example, it would appear that
    the mean age of the wives is just a shade under 1500 years.

-   The `replace_with_na` function from the `naniar` package (as it is not
    already loaded via `library`, we utilise this function by prefacing
    with the package name and two colons) replaces the entries given in
    the `replace` list with `NA`. In this case we've asked to replace 9999
    in both the `W.Age` and `H.Age.Marriage` columns. We've saved the result
    into the `husb_fixed` tibble.

-   Applying the function `mean` to a variable containing missing values
    will by default return `NA` as a result. However, if we set the
    optional argument `na.rm=TRUE` (i.e. remove missing values) then we get
    the mean for the remaining data. In the case of wives' ages, that is
    just over 40.
    
In this example, as the special value that denotes missingness is consistent
across all columns, we could use the `na` argument of `read_csv`. It defaults
to `c("", "NA")`, to detect the empty string or the string NA, but we could
replace this with `"9999"`:

```{r, message=FALSE, render = knitr::normal_print}
husbands <- read_csv(file="../data/husbands.csv", na="9999")
husbands %>% skimr::skim()
```

:::

[^otherpackages]: R can also interact with database systems as well, although the
    process is a little more complex.

[^factors]: As noted previously, categorical variables in R are typically
    referred to as factors. The different categories themselves are
    termed the *levels* of the factor. Hence the factor `Treatment` in
    the goats dataset has two levels, `standard` and `intensive`
    distinguishing the drenching routine employed.

[^tibble]: There are some subtle differences between tibbles and data.frames. Hopefully
we won't hit too many of them!

[^droppingout]: An important example occurs with data from clinical trials, where
    the statistical analysis is typically expected to account for
    *intention to treat*. This prevents a misleadingly statistical
    results in cases where subjects who are responding poorly have a
    disproportionate chance of dropping out of the trial.

[^nasodium]: You might run into problems if the data include a list of chemical
    elements, for example! The chemical symbol for sodium is Na, from
    the Latin word *natrium*.
